---
title:  "2024-25 NBA Salary Data"
format:
  closeread-html:
    cr-style:
      section-background-color: "white"
      narrative-background-color-sidebar: "#1D428A"
      narrative-text-color-sidebar: white
      embed-resources: true
---

```{r, include=FALSE}
library(tidyverse)
library(plotly)
library(knitr)
options(scipen=999)
```
```{r, include=FALSE}
nba = read_csv("nba2025-clean.csv")
```

:::: {.cr-section}

![](https://cdn.nba.com/logos/leagues/logo-nba.svg){width=50%}

Salary Distribution and Statistics @cr-summary

::: {#cr-summary}
```{r}
ggplotly(ggplot(data=nba, aes(x=salary)) + geom_histogram(bins=50) + labs(x="Salary", y="Count"))
kable(summarize(nba, Min = min(salary),
    Q1 = quantile(salary, 0.25),
    Median = median(salary),
    Mean = mean(salary),
    Q3 = quantile(salary, 0.75),
    Max = max(salary)))
```
:::

10 Highest Salaries @cr-t10

Among star players, we see multiple contracts paying $49.2M in 2024-25 as a result of the salary cap and the max contract value being limited to 35% of the cap for players with 10+ years of experience. @cr-t10g

look at curry man so inspirational @cr-steph

::: {#cr-t10}
```{r}
t10 = nba |> select("player","team","position","salary") |> arrange(desc(salary)) |> slice_head(n=10)
plot = ggplot(data=t10, aes(x=player,y=salary)) + geom_col() + coord_cartesian(ylim=c(45000000,60000000)) + theme(axis.text.x = element_text(angle = 315)) + labs(x="Player",y="Salary")
ggplotly(plot)
kable(t10)
```
:::

::: {#cr-t10g}
```{r}
plot2 = ggplot(data=t10, aes(x=player,y=salary,fill=salary<50000000)) + geom_col() + coord_cartesian(ylim=c(45000000,60000000)) + theme(axis.text.x = element_text(angle = 315),legend.position="none") + labs(x="Player",y="Salary") + scale_fill_manual(values = c("TRUE"="green"))
ggplotly(plot2)
kable(t10)
```
:::

::: {#cr-steph}
```{r}
plot5 = ggplot(data=t10, aes(x=player,y=salary,fill=salary>55000000)) + geom_col() + coord_cartesian(ylim=c(45000000,60000000)) + theme(axis.text.x = element_text(angle = 315),legend.position="none") + labs(x="Player",y="Salary") + scale_fill_manual(values = c("TRUE"="#1D428A"))
ggplotly(plot5) |> plotly::layout(autosize=TRUE) |> plotly::config(responsive=TRUE)
kable(t10)
```
:::

10 Highest Team Payrolls @cr-payroll

Bradley Beal is singlehandedly responsible for the Suns payroll. Book and KD have always earned their share; Beal pulled off one of the greatest contract heists of the decade. @cr-beal

:::{#cr-payroll}
```{r}
tpayroll = nba |> group_by(team) |> summarize(payroll=sum(salary)) |> arrange(desc(payroll)) |> slice_head(n=10)
plot3 = ggplot(data=tpayroll, aes(x=team,y=payroll)) + geom_col() + coord_cartesian(ylim=c(150000000,250000000)) + labs(x="Team",y="Payrooll")
ggplotly(plot3)
```
:::

:::{#cr-beal}
```{r}
plot4 = ggplot(data=tpayroll, aes(x=team,y=payroll,fill=payroll>200000000)) + geom_col() + coord_cartesian(ylim=c(150000000,250000000)) + theme(legend.position="none") + labs(x="Team",y="Payroll") + scale_fill_manual(values = c("TRUE"="#E56020"))
ggplotly(plot4)
```
:::

Salary vs Experience @cr-xp

We see a clear salary limit in the first four years of careers. First-round picks are limited to fixed-scale 4-year rookie contracts, while later picks are typically only offered contracts at or near the minimum. @cr-rookie

These restrictions are substantially lifted after the rookie contract expires, or from the 4th year onward, in the case of rookie-scale extensions. @cr-xp

:::{#cr-xp}

```{r}
ggplotly(ggplot(nba, aes(x=experience, y=salary)) + geom_point(alpha=0.3) + labs(x="Experience", y="Salary"))
```

:::

:::{#cr-rookie}

```{r}
ggplotly(ggplot(nba, aes(x=experience, y=salary)) + geom_point(alpha=0.3) + annotate("segment",x = 0, xend = 3, y = 16000000, yend = 16000000, color="red") + labs(x="Experience", y="Salary"))
```

:::

Salary Market by Position @cr-pos

Teams tend to value great point guards more than great players in other positions, like steph. @cr-pg

Look at bron at the top of the sf market. @cr-sf

:::{#cr-pos}

```{r}
ggplotly(ggplot(nba, aes(x=position, y=salary)) + geom_violin() + labs(x="Position", y="Salary"))
```
:::

:::{#cr-pg}

```{r}
ggplotly(ggplot(nba, aes(x=position, y=salary, fill=position=="PG")) + geom_violin() + labs(x="Position", y="Salary") + scale_fill_manual(values = c("TRUE" = "#1D428A", "FALSE" = "white")) + theme(legend.position="none"))
```
:::

:::{#cr-sf}

```{r}
ggplotly(ggplot(nba, aes(x=position, y=salary, fill=position=="SF")) + geom_violin() + labs(x="Position", y="Salary") + scale_fill_manual(values = c("TRUE" = "#860038", "FALSE" = "white")) + theme(legend.position="none"))
```
:::

go dubs @cr-pos

::::